---
# YAML header created by ox-ravel
title: Non-linear latent variable models and error-in-variable models
author: Klaus Kähler Holst
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Non-linear latent variable models and error-in-variable models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<!-- nonlinear.Rmd is generated from nonlinear.org. Please edit that file -->

```{r  include=FALSE }
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
mets <- lava:::versioncheck('mets', 1)
fullVignette <- Sys.getenv("_R_FULL_VIGNETTE_") %in% c("1","TRUE")
```

```{r  load, results="hide",message=FALSE,warning=FALSE }
library('lava')
```

We consider the measurement models given by

\[X_{j} = \eta_{1} + \epsilon_{j}^{x}, \quad j=1,2,3\]
\[Y_{j} = \eta_{2} + \epsilon_{j}^{y}, \quad j=1,2,3\]
and with a structural model given by
\[\eta_{2} = f(\eta_{1}) + Z + \zeta_{2}\label{ex:eta2}\]
\[\eta_{1} = Z + \zeta_{1}\label{ex:eta1}\]
with iid measurement errors
\(\epsilon_{j}^{x},\epsilon_{j}^{y},\zeta_{1},\zeta_{2}\sim\mathcal{N}(0,1),
j=1,2,3.\) and standard normal distributed covariate \(Z\).  To
simulate from this model we use the following syntax:

```{r  sim }
f <- function(x) cos(1.25*x) + x - 0.25*x^2
m <- lvm(x1+x2+x3 ~ eta1, y1+y2+y3 ~ eta2, latent=~eta1+eta2)
regression(m) <- eta1+eta2 ~ z
functional(m, eta2~eta1) <- f

d <- sim(m, n=200, seed=42) # Default is all parameters are 1
```

```{r   }
plot(m, plot.engine="visNetwork")
```

We refer to <sup id="915607a5e4b7965311f3b3d4e0ccb7d7"><a href="#holst_budtzjorgensen_2013" title="Holst \&amp; Budtz-J\orgensen, Linear latent variable models: the lava-package, {Computational Statistics}, v(4), 1385--1452 (2013).">holst_budtzjorgensen_2013</a></sup> for details on the syntax for
model specification.


# Estimation

To estimate the parameters using the two-stage estimator described in <sup id="67e17d4cdc94644cb044c767cf392c2d"><a href="#holst_budtzjorgensen_2020" title="Klaus K&#228;hler Holst \&amp; Esben Budtz-J&#248;rgensen, A two-stage estimation procedure for non-linear structural equation models, {Biostatistics}, v(), (2020).">holst_budtzjorgensen_2020</a></sup>,  the first step is now to specify the measurement models

```{r  specifymodels }
m1 <- lvm(x1+x2+x3 ~ eta1, eta1 ~ z, latent=~eta1)
m2 <- lvm(y1+y2+y3 ~ eta2, eta2 ~ z, latent=~eta2)
```

Next, we specify a quadratic relationship between the two latent variables

```{r   }
nonlinear(m2, type="quadratic") <- eta2 ~ eta1
```

and the model can then be estimated using the two-stage estimator

```{r  twostage1 }
e1 <- twostage(m1, m2, data=d)
e1
```

We see a clear statistically significant effect of the second order
term (`eta2~eta1_2`). For comparison we can also estimate the full MLE
of the linear model:

```{r  linear_mle }
e0 <- estimate(regression(m1%++%m2, eta2~eta1), d)
estimate(e0,keep="^eta2~[a-z]",regex=TRUE) ## Extract coef. matching reg.ex.
```

Next, we calculate predictions from the quadratic model using the estimated parameter coefficients
\[
\mathbb{E}_{\widehat{\theta}_{2}}(\eta_{2} \mid \eta_{1}, Z=0),
\]

```{r  pred1 }
newd <- expand.grid(eta1=seq(-4, 4, by=0.1), z=0)
pred1 <- predict(e1, newdata=newd, x=TRUE)
head(pred1)
```

To obtain a potential better fit we next proceed with a natural cubic spline

```{r  spline_twostage }
kn <- seq(-3,3,length.out=5)
nonlinear(m2, type="spline", knots=kn) <- eta2 ~ eta1
e2 <- twostage(m1, m2, data=d)
e2
```

Confidence limits can be obtained via the Delta method using the `estimate` method:

```{r  spline_ci }
p <- cbind(eta1=newd$eta1,
	  estimate(e2,f=function(p) predict(e2,p=p,newdata=newd))$coefmat)
head(p)
```

The fitted function can be obtained with the following code:

```{r  fig:pred2 }
plot(I(eta2-z) ~ eta1, data=d, col=Col("black",0.5), pch=16,
     xlab=expression(eta[1]), ylab=expression(eta[2]), xlim=c(-4,4))
lines(Estimate ~ eta1, data=as.data.frame(p), col="darkblue", lwd=5)
confband(p[,1], lower=p[,4], upper=p[,5], polygon=TRUE,
	 border=NA, col=Col("darkblue",0.2))
```


# Cross-validation

A more formal comparison of the different models can be obtained by
cross-validation. Here we specify linear, quadratic and cubic spline
models with 4 and 9 degrees of freedom.

```{r  spline_several }
m2a <- nonlinear(m2, type="linear", eta2~eta1)
m2b <- nonlinear(m2, type="quadratic", eta2~eta1)
kn1 <- seq(-3,3,length.out=5)
kn2 <- seq(-3,3,length.out=8)
m2c <- nonlinear(m2, type="spline", knots=kn1, eta2~eta1)
m2d <- nonlinear(m2, type="spline", knots=kn2, eta2~eta1)
```

To assess the model fit average RMSE is estimated with 5-fold
cross-validation repeated two times

```{r  cv_fit, cache=TRUE, eval=fullVignette }
## Scale models in stage 2 to allow for a fair RMSE comparison
d0 <- d
for (i in endogenous(m2))
    d0[,i] <- scale(d0[,i],center=TRUE,scale=TRUE)
## Repeated 5-fold cross-validation:
ff <- lapply(list(linear=m2a,quadratic=m2b,spline4=m2c,spline6=m2d),
	    function(m) function(data,...) twostage(m1,m,data=data,stderr=FALSE,control=list(start=coef(e0),contrain=TRUE)))
fit.cv <- cv(ff,data=d,K=5,rep=2,mc.cores=parallel::detectCores(),seed=1)
```

```{r  results="hide", echo=FALSE }
## To save time building the vignettes on CRAN, we cache time consuming computations
if (fullVignette) {
  fit.cv$fit <- NULL
  saveRDS(fit.cv, "data/nonlinear_fitcv.rds")
} else {
  fit.cv <- readRDS("data/nonlinear_fitcv.rds")
}
```

```{r   }
summary(fit.cv)
```

Here the RMSE is in favour of the splines model with 4 degrees of freedom:

```{r  multifit }
fit <- lapply(list(m2a,m2b,m2c,m2d),
	     function(x) {
		 e <- twostage(m1,x,data=d)
		 pr <- cbind(eta1=newd$eta1,predict(e,newdata=newd$eta1,x=TRUE))
		 return(list(estimate=e,predict=as.data.frame(pr)))
	     })

plot(I(eta2-z) ~ eta1, data=d, col=Col("black",0.5), pch=16,
     xlab=expression(eta[1]), ylab=expression(eta[2]), xlim=c(-4,4))
col <- c("orange","darkred","darkgreen","darkblue")
lty <- c(3,4,1,5)
for (i in seq_along(fit)) {
    with(fit[[i]]$pr, lines(eta2 ~ eta1, col=col[i], lwd=4, lty=lty[i]))
}
legend("bottomright",
      c("linear","quadratic","spline(df=4)","spline(df=6)"),
      col=col, lty=lty, lwd=3)
```

For convenience, the function `twostageCV` can be used to do the
cross-validation (also for choosing the mixture distribution via the \`\`nmix\`\` argument, see the section
below). For example,

```{r  twostageCV, cache=TRUE, eval=fullVignette }
selmod <- twostageCV(m1, m2, data=d, df=2:4, nmix=1:2,
	    nfolds=2, rep=1, mc.cores=parallel::detectCores())
```

```{r  results="hide", echo=FALSE }
## To save time building the vignettes on CRAN, we cache time consuming computations
if (fullVignette) {
  saveRDS(summary(selmod), "data/nonlinear_selmod.rds")
} else {
  selmod <- readRDS("data/nonlinear_selmod.rds")
}
```

applies cross-validation (here just 2 folds for simplicity) to select the best splines with
degrees of freedom varying from from 1-3 (the linear model is
automatically included)

```{r   }
selmod
```


# Specification of general functional forms

Next, we show how to specify a general functional relation of
multiple different latent or exogenous variables. This is achieved via
the `predict.fun` argument. To illustrate this we include interactions
between the latent variable \(\eta_{1}\) and a dichotomized version of
the covariate \(z\)

```{r   }
d$g <- (d$z<0)*1 ## Group variable
mm1 <- regression(m1, ~g)  # Add grouping variable as exogenous variable (effect specified via 'predict.fun')
mm2 <- regression(m2, eta2~ u1+u2+u1:g+u2:g+z)
pred <- function(mu,var,data,...) {
    cbind("u1"=mu[,1],"u2"=mu[,1]^2+var[1],
	  "u1:g"=mu[,1]*data[,"g"],"u2:g"=(mu[,1]^2+var[1])*data[,"g"])
}
ee1 <- twostage(mm1, model2=mm2, data=d, predict.fun=pred)
estimate(ee1,keep="eta2~u",regex=TRUE)
```

A formal test show no statistically significant effect of this interaction

```{r   }
summary(estimate(ee1,keep="(:g)", regex=TRUE))
```


# Mixture models

Lastly, we demonstrate how the distributional assumptions of stage 1
model can be relaxed by letting the conditional distribution of the
latent variable given covariates follow a Gaussian mixture
distribution. The following code explictly defines the parameter
constraints of the model by setting the intercept of the first
indicator variable, \(x_{1}\), to zero and the factor loading
parameter of the same variable to one.

```{r   }
m1 <- baptize(m1)  ## Label all parameters
intercept(m1, ~x1+eta1) <- list(0,NA) ## Set intercept of x1 to zero. Remove the label of eta1
regression(m1,x1~eta1) <- 1 ## Factor loading fixed to 1
```

The mixture model may then be estimated using the `mixture` method
(note, this requires the `mets` package to be installed), where the
Parameter names shared across the different mixture components given
in the `list` will be constrained to be identical in the mixture
model. Thus, only the intercept of \(\eta_{1}\) is allowed to vary
between the mixtures.

```{r  mixture1, cache=TRUE, eval=fullVignette }
set.seed(1)
em0 <- mixture(m1, k=2, data=d)
```

To decrease the risk of using a local maximizer of the likelihood we
can rerun the estimation with different random starting values

```{r  estmixture, cache=TRUE,warnings=FALSE,messages=FALSE,eval=FALSE }
em0 <- NULL
ll <- c()
for (i in 1:5) {
    set.seed(i)
    em <- mixture(m1, k=2, data=d, control=list(trace=0))
    ll <- c(ll,logLik(em))
    if (is.null(em0) || logLik(em0)<tail(ll,1))
	em0 <- em
}
```

```{r  results="hide", echo=FALSE }
## To save time building the vignettes on CRAN, we cache time consuming computations
if (fullVignette) {
  saveRDS(em0, "data/nonlinear_em0.rds")
} else {
  em0 <- readRDS("data/nonlinear_em0.rds")
}
```

```{r   }
summary(em0)
```

Measured by AIC there is a slight improvement in the model fit using the mixture model

```{r  eval=mets }
e0 <- estimate(m1,data=d)
AIC(e0,em0)
```

The spline model may then be estimated as before with the `two-stage` method

```{r  eval=mets }
em2 <- twostage(em0,m2,data=d)
em2
```

In this example the results are very similar to the Gaussian model:

```{r  mixturefit, eval=mets }
plot(I(eta2-z) ~ eta1, data=d, col=Col("black",0.5), pch=16,
     xlab=expression(eta[1]), ylab=expression(eta[2]))

lines(Estimate ~ eta1, data=as.data.frame(p), col="darkblue", lwd=5)
confband(p[,1], lower=p[,4], upper=p[,5], polygon=TRUE,
	 border=NA, col=Col("darkblue",0.2))

pm <- cbind(eta1=newd$eta1,
	    estimate(em2, f=function(p) predict(e2,p=p,newdata=newd))$coefmat)
lines(Estimate ~ eta1, data=as.data.frame(pm), col="darkred", lwd=5)
confband(pm[,1], lower=pm[,4], upper=pm[,5], polygon=TRUE,
	 border=NA, col=Col("darkred",0.2))
legend("bottomright", c("Gaussian","Mixture"),
       col=c("darkblue","darkred"), lwd=2, bty="n")
```


# Bibliography
<a id="holst_budtzjorgensen_2013"></a>[holst_budtzjorgensen_2013] Holst & Budtz-J\orgensen, Linear latent variable models: the lava-package, <i>Computational Statistics</i>, <b>28(4)</b>, 1385-1452 (2013). <a href="http://dx.doi.org/10.1007/s00180-012-0344-y">doi</a>. [↩](#915607a5e4b7965311f3b3d4e0ccb7d7)

<a id="holst_budtzjorgensen_2020"></a>[holst_budtzjorgensen_2020] Klaus Kähler Holst & Esben Budtz-Jørgensen, A two-stage estimation procedure for non-linear structural equation models, <i>Biostatistics</i>, <b>(in press)</b>, (2020). <a href="http://dx.doi.org/10.1093/biostatistics/kxy082">doi</a>. [↩](#67e17d4cdc94644cb044c767cf392c2d)
